Beyond the best guess: Distributional predictions from large language models for human forced-choice data

### 2023-10-06 03-paper/paper-LLM-probmod.tex
revised §2

The prevalent approach to assessing LLM performance based on benchmark-testing quantifies accuracy for a multiple-choice task with a ``winner-takes-all'' (WTA) strategy.
Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances of the same task, or items belonging to the same (logical) condition in a behavioral experiment.
% For human observers, these items are interchangeable, all exemplifying the same conceptual problem.
Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice alternatives $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ words, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$.
The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
  More elaborate item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
  From the point of view of experimental psychology, these corrections are \emph{post hoc} fixes to improperly balanced experimental materials.
  For the purposes of this paper, where answer options are all equally long and commensurable, these corrections may be temporarily ignored for simplicity.
}
%
\begin{align*}
  S_{\text{item}}\left( y_{ki}, x_{k} \right) = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right) =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
\end{align*}
%
The WTA approach considers item $I_{k}$ to be answered correctly if the designated ``goal answer'' $y_{ki^{*}}$ is among the options that maximize the item-level score.
The accuracy of the LLM on the given task is the proportion of items in which the correct answer is selected. \mf{Rephrase or add FN on tie-breaking.}

While generally a pragmatic and useful approach, there are several reasons why going beyond the best guess can be beneficial.
Consider first a high-level conceptual argument.
If task performance is categorical (in the most extreme case: binary), somewhere along the path from a numerical item-level score to accuracy a discontinuity has to be introduced.
Every such discontinuity bottleneck entails loss of information.
As this information might be useful, discontinuity should ideally happen as late as possible; or even better: not at all, unless it is irrelevant.
Consequently, the question becomes: when is the additional information inherent in the scores (not) relevant for assessing LLM performance?

The WTA approach implicitly assumes that item-level choices are resolved by a greedy-like choice of the best alternative.
It is therefore a good measure of accuracy when the LLM's main use cases are of a greedy-like nature as well.
When applications rely on a less extreme sampling based strategy, the WTA approach can be generalized to softmax sampling.
The probability of choosing option $y_{ki}$ for item $I_{k}$ is:
%
\begin{align*}
P_{\text{choose}}(y_{ki}) \propto \expo \left [ \alpha \ S_{\text{item}}\left( y_{ki}, x_{k} \right) \right ]\,,
\end{align*}
%
and the accuracy is the average probability of choosing the designated target option $y_{ki^{*}}$.
For $\alpha \rightarrow \infty$, this softmax strategy converges to the WTA strategy.
But in general, these two approaches can yield even categorically different predictions.

Imagine that there are two options, and that the target option is a small $\epsilon$ better in 80\% of the task's items, and otherwise worse.
The WTA-based accuracy is 0.8.
This number is useful as a performance measure for applications in which the LLM is used in exactly the way the WTA strategy describes, e.g., any implementation which is outcome equivalent to greedy decoding with rejection sampling on a domain that contains only the available options.
For such a case, it never matters how much worse the goal answer is scored in the 20\% of the cases where it is not the maximum.
As only the best option will be chosen, that information is irrelevant.
But if an application uses anything other than greedy-like responses, the accuracy score of 0.8 may be misleading.
If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy, where $\alpha = 1$, with virtual certainty, so the softmax-based accuracy would around 0.4.\footnote{The probability of the second option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 1 in the remaining 20\%.}
This is a categorical shift from predominantly goal answer to predominantly non-goal answer.
The argument holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the goal answers.

Differences in accuracy scores between a WTA-based or a more general softmax-based strategy depend on the properties of the variation in item-level scores, in particular the relation between score-ordering and score-differences.
(The example was set up deliberately to have one score-ordering for very small score-differences, and another for huge score-differences.)
This raises an important general question, namely how to deal with item-level variation in a system's predictions in the first place.
Data from forced-choice task with human participants will likely show item-level variation, which is why hierarchical modeling accommodating item-level differences is recommended in statistical analyses \citep[e.g.,][]{Jaeger2008:Categorical-dat,barr2013}.
But whether item-level variation in LLM scores aligns with differences in human choice rates for each item is an empirical question that has to be addressed by explicit modeling and comparison; an exercise that this paper starts to take on.
On \emph{a priori} grounds, the WTA-based strategy may be deemed implausible as a predictor of human-level choice behavior for a simple item: it is unlikely that, for each item, there is a single choice option that (almost) \emph{all} participants will choose.

@@

\mf{To address whether LLM's condition-level predictions are adequate as item-level predictors, I should re-run the item-level analysis with the mean posterior predictive of the intermediate-scope model (instead of, or in addition to RSA)!}

>>>>

The prevalent approach to assessing LLM performance is based on benchmark-testing, using large data sets for tasks in which a designated goal answer exists.
Tasks often contain multiple-choice options, and accuracy for a task is determined with a ``winner-takes-all'' (WTA) strategy \citep[e.g.,][]{srivastava2023-BIGbench}.

Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances of the same task, or items belonging to the same (logical) condition in a behavioral experiment.
% For human observers, these items are interchangeable, all exemplifying the same conceptual problem.
Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice options $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ words, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$.
Among the choice options is a designated \emph{target option} $y_{ki^{*}}$ that is assumed to be the ``true'' or to-be-selected goal answer.

The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
  More elaborate item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
  From the point of view of experimental psychology, these corrections are \emph{post hoc} fixes to improperly balanced experimental materials.
  For the purposes of this paper, where answer options are all equally long and commensurable, these corrections may be temporarily ignored for simplicity.
}
%
\begin{align*}
  S\left( y_{ki}, x_{k} \right) = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right) =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
\end{align*}
%
The WTA approach considers item $I_{k}$ to be answered correctly if the target option $y_{ki^{*}}$ is the option that maximizes the item-level score.
The WTA-based accuracy of the LLM on the given task is the proportion of items which are answered correctly.

\mf{the following paragraphs should be shortened and sharpened}
While generally a pragmatic and useful approach, there are several reasons why going beyond the best guess can be beneficial.
Consider first a high-level conceptual argument.
If task performance is categorical (in the most extreme case: binary), somewhere along the path from a numerical item-level score to accuracy a discontinuity has to be introduced.
Every such discontinuity bottleneck entails loss of information.
As this information might be useful, discontinuity should ideally happen as late as possible; or even better: not at all, unless it is irrelevant.
Consequently, the question becomes: when is the additional information inherent in the scores (not) relevant for assessing LLM performance?

The WTA approach implicitly assumes that item-level choices are resolved by a greedy-like choice of the best alternative.
It is therefore a good measure of accuracy when the LLM's main use cases are of a greedy-like nature as well.
When applications rely on a less extreme sampling based strategy, the WTA approach can be generalized to softmax sampling.
The probability of choosing option $y_{ki}$ for item $I_{k}$ is:
%
\begin{align*}
P_{\text{choose}}(y_{ki}) \propto \expo \left [ \alpha \ S \left( y_{ki}, x_{k} \right) \right ]\,,
\end{align*}
%
and the accuracy is the average probability of choosing the designated target option $y_{ki^{*}}$.
For $\alpha \rightarrow \infty$, this softmax strategy converges to the WTA strategy.
But in general, these two approaches can yield even categorically different predictions.

Imagine that there are two options, and that the target option's score is a small $\epsilon$ better in 80\% of the task's items, and otherwise worse.
The WTA-based accuracy is 0.8.
This number is useful as a performance measure for applications in which the LLM is used in exactly the way the WTA strategy describes, e.g., any implementation which is outcome equivalent to greedy decoding with rejection sampling on a domain that contains only the available options.
For such a case, it never matters how much worse the goal answer is scored in the 20\% of the cases where it is not the maximum.
As only the best option will be chosen, that information is irrelevant.
But if an application uses anything other than greedy-like responses, the accuracy score of 0.8 may be misleading.
If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy, where $\alpha = 1$, with virtual certainty, so the softmax-based accuracy would around 0.4.\footnote{The probability of the target option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 0 in the remaining 20\% of the cases. This gives an expected rate of: $\nicefrac{4}{5} \ \nicefrac{1}{2} + \nicefrac{1}{5} \ 0 = \nicefrac{2}{5}$.}
This is a categorical shift from predominantly goal answer to predominantly non-goal answer.
The argument holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.

Differences in accuracy scores between a WTA-based or a more general softmax-based strategy depend on the variation in item-level scores, in particular the relation between score-ordering and score-differences.
(The example in the previous paragraph was set up deliberately to have one score-ordering for very small score-differences, and another for huge score-differences.)
This raises an important general question, namely how to deal with item-level variation in a system's predictions in the first place.
Data from forced-choice tasks with human participants will likely show item-level variation as well, which is why hierarchical modeling accommodating item-level differences is recommended in statistical analyses \citep[e.g.,][]{Jaeger2008:Categorical-dat,barr2013}.
But whether item-level variation in LLM scores aligns with differences in human choice rates for each item is an empirical question that has to be addressed by explicit modeling and comparison; an exercise that this paper starts to take on.
On \emph{a priori} grounds, the WTA-based strategy may be deemed implausible as a predictor of human-level choice behavior for a simple item: it is unlikely that, for each item, there is a single choice option that (almost) \emph{all} participants will choose.

@@
